---
title: "FFG HDAC HAP RNA-seq: Sample similarities"
author: "Adrian Arnel Lauber"
date: "18.03.2021"
output:
  html_document: 
    theme: cerulean
    highlight: monochrome
    toc: yes
    toc_float: yes
  word_document:
    toc: yes
  pdf_document: 
    toc: yes
  html_notebook: 
    toc: yes
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE,
                      include = TRUE,
                      warning=FALSE,
                      message=TRUE,
                      fig.align = 'center')
```

## Load libraries and initialize parallelication according to OS
```{r imports}
library("tidyverse")
library("DESeq2")
library("pheatmap")
```

```{r parallelization}
platform <- Sys.info()["sysname"]

if (platform == "Linux") {
  library("BiocParallel")
  register(MulticoreParam(2))
}
```

# Background & Summary
The samples from the inhibitor model design are analysed for deregulated genes. Samples MS275, JQ12, ...

## Read raw count matrix and coldata + extract metadata
  * Raw count matrix is provided by Michael Schuster of CEMM. 
  * nome: Homo sapiens genome assembly hg38 (December 2013)
  * Transcriptome: Homo sapiens transcriptome hg38_e100 (2020?)
  * STAR aligner was used for mapping of reads.
  * The same RAW count numbers which where used in Schuster´s DEseq analysis are being read in here
  * Coldata: Additional information about each sample. they are stored in the Metadata slot of the DEseq object


The "inhibitor_counts_raw.tsv" includes all samples which were used in the finale analysis. The samples of HDAC3 which were not True Knock-outs are not in this table anymore and HDAC3 Knock-in with the mutant Transgene is included.
I was not sure whether the samples with the wt HDAC TG in KO background is still in the table, that's why I excluded it here.
Since the countmatrix input for DESeq needs only the gene_id and the raw counts, the columns with additional information is extracted (featureData) and passed below to the DESeqDataSet object into mcols.

The specific sample information with factors is extracted from the "..._samples.tsv", unnecessary columns are dropped.

```{r countmatrix}
cts_full <- read_tsv("rnaseq_deseq_inhibitor_counts_raw.tsv")
cts <- cts_full %>% select(gene_id, starts_with("H") & !contains("_wt_")) %>% column_to_rownames("gene_id")
cts_matrix <- as.matrix(cts)

coldata <- read.csv("rnaseq_deseq_inhibitor_samples.tsv", sep="\t", row.names=12, colClasses=c("NULL","NULL", "NULL", NA, NA, "NULL", NA, NA, NA, NA, "NULL", NA, "NULL", NA), stringsAsFactors = TRUE)

featureData <- cts_full %>% select(gene_version:location)

```

Difference between sample data and count matrix samples?
Then remove these rows and check whether sample columns and rows fit

```{r difference}
ctsc <- colnames(cts_matrix)
coldatac <- rownames(coldata)
(setdiff(coldatac, ctsc))

coldata <- coldata[!(row.names(coldata) %in% setdiff(rownames(coldata),ctsc)), ]
(all(rownames(coldata) == colnames(cts)))
```

## Create dds object + add metadata
As soon as the sample data from coldata matches the samples from the count Matrix, the dds Object can be initialized and our Metadata (featureData) can also be added so we can have everything in the dds object

```{r dds_object}
dds <- DESeqDataSetFromMatrix(countData = cts_matrix,
                              colData = coldata,
                              design = ~ group)

mcols(dds) <- DataFrame(mcols(dds), featureData)
```

## relevel factors to set the reference

```{r relevel}
dds$group <- relevel(dds$group, ref = "wt_native")
```

## Pre-filter the low raw count rows
Here we can define the threshold for the sum of raw counts per row, so that rows (=genes) with a generally low count can be dropped. This threshold can also depend on the number of samples!

```{r prefilter}
keep <- rowSums(counts(dds)) > 1
dds <- dds[keep,]
```

## make a copy of standard DESeq (wald) dds object for lrt
The lrt analysis can be used to make a statistical analysis over all samples instead just pairwise comparisons.

```{r copy_for_lrt}
dds_lrt <- dds
```

## Deseq with LRT

```{r deseq_lrt}
if (platform == "Linux") {
  dds_lrt <- DESeq(dds_lrt, test="LRT", reduced=~1, parallel=TRUE)
} else {
  dds_lrt <- DESeq(dds_lrt, test="LRT", reduced=~1)
}

res_lrt <- results(dds_lrt)
```

## Adding gene names to lrt test results of DEseq

```{r add_gene_names_lrt}
res_lrt$symbol <- mcols(dds_lrt)$gene_name
res_lrt$gene_biotype <- mcols(dds_lrt)$gene_biotype
```

## Analyse and export results (lrt)

```{r results_lrt}

res_lrt_sig <- subset(res_lrt, padj < 0.1)
write.csv(as.data.frame(res_lrt), 
          file="20210318_inhibitor_all/inhibitor_wo_group_results.csv")
write.csv(as.data.frame(res_lrt_sig), 
          file="20210318_inhibitor_all/inhibitor_wo_group_results_significant.csv")
```


## Deseq standard (wald test)

```{r deseq_wald}
if (platform == "Linux") {
  dds <- DESeq(dds, parallel=TRUE)
} else {
  dds <- DESeq(dds)
}

res <- results(dds)
```

## Adding gene names and gene types

```{r add_gene_names}
res$gene_name <- mcols(dds)$gene_name
res$gene_biotype <- mcols(dds)$gene_biotype
```

## Quality and outlier control
Here just MA plots to look at the distribution. Scipen Option will stop R to write the numbers on the X-Axis in scientific notation.

```{r deseq_control}
summary(res)
options(scipen = 5)

plotMA(res, ylim = c(-10,10), main = "group wt MS275 3uM 24h vs wt native \n rowsum > 1: 38311 genes")

topGene <- rownames(res)[which.min(res$padj)]
with(res[topGene, ], {
  points(baseMean, log2FoldChange, col = "red", cex = 2, lwd = 2)
  text(baseMean, log2FoldChange, topGene, pos = 2, col = "red")
})

options(scipen = 0)
```

## Analyse and export results (standard wald test)
padj was defined 2x since I looked at the total genes and the ordered MS275 list seperately... This is not clear...

```{r results_wald}

MS275_3uM_ordered <- res[order(res$log2FoldChange),]
MS275_3uM_ordered_sig <- subset(MS275_3uM_ordered, padj < 0.1)
write.csv(as.data.frame(MS275_3uM_ordered_sig), 
          file="20210318_inhibitor_all/group_wt_MS275_3uM_24h_vs_wt_native.csv")

jq12 <- results(dds, contrast=c("group", "wt_JQ12_1uM_24h", "wt_native"))
jq12_ordered <- jq12[order(jq12$log2FoldChange),]
jq12_ordered_sig <- subset(jq12_ordered, padj < 0.1)
write.csv(as.data.frame(jq12_ordered_sig), 
          file="20210318_inhibitor_all/group_wt_JQ12_1uM_24h_vs_wt_native.csv")

```

## export all contrasts versus wt_native

```{r export_contrasts}
all_contr <- resultsNames(dds)

for (val in all_contr)
{
if(is.character(val) == TRUE) 
  act_contr <- results(dds, name=val)
  act_contr$gene_name <- mcols(dds)$gene_name
  act_contr$gene_biotype <- mcols(dds)$gene_biotype
  act_contr_ordered <- act_contr[order(act_contr$log2FoldChange),]
  act_contr_ordered_sig <- subset(act_contr_ordered, padj < 0.1)
  write.csv(as.data.frame(act_contr_ordered_sig), 
          file=paste0("20210318_inhibitor_all/", val, ".csv"))
}

```

## create gene symbol lists for further analysis of gene set overlaps
The contrasts which compare the genes to the wt_native are extracted from "results" and ordered according to the fold change. Then only significant genes with an adjusted p-value of < 0.1 and a log2FoldChange bigger than 1  are chosen. First upregulated genes and then in second loop downregulated genes.

```{r comparison_lists}
all_contr <- resultsNames(dds)

for (val in all_contr)
{
if(is.character(val) == TRUE) 
  act_contr <- results(dds, name=val)
  act_contr$gene_name <- mcols(dds)$gene_name
  act_contr$gene_biotype <- mcols(dds)$gene_biotype
  act_contr_ordered <- act_contr[order(act_contr$log2FoldChange, decreasing=TRUE),]
  act_contr_ordered_sig <- subset(act_contr_ordered, padj < 0.1 & log2FoldChange > 1)
  write.csv(as.data.frame(act_contr_ordered_sig$gene_name), 
          file=paste0("20210318_inhibitor_all/symb_list/", val, "_up.csv"), row.names = FALSE)
}

for (val in all_contr)
{
if(is.character(val) == TRUE) 
  act_contr <- results(dds, name=val)
  act_contr$gene_name <- mcols(dds)$gene_name
  act_contr_ordered <- act_contr[order(act_contr$log2FoldChange, decreasing=TRUE),]
  act_contr_ordered_sig <- subset(act_contr_ordered, padj < 0.1 & log2FoldChange < -1)
  write.csv(as.data.frame(act_contr_ordered_sig$gene_name), 
          file=paste0("20210318_inhibitor_all/symb_list/", val, "_down.csv"), row.names = FALSE)
}
```

## TODO: This part is for extraction of HDACs and plot to see change of isoforms in cell lines
This part is still ongoing, datasets have to be swapped

```{r expression_levels_groups, eval=FALSE, include=FALSE}
data1 <- plotCounts(dds, "ENSG00000116478",
                   intgroup = c("type", "cell"), returnData = TRUE)
data1 <- data1 %>% mutate(cell = replace_na(cell, "HDAC1")) %>% filter(type != "wt_jq12") %>% filter(!str_detect(type, "^wt_ms275"))
ggplot(data1, aes(x=type, y=count, color = cell, group = cell)) +
  geom_point() + stat_smooth(se = FALSE, method = "loess") + scale_y_log10()

data2 <- plotCounts(dds, "ENSG00000196591",
                   intgroup = c("type", "cell"), returnData = TRUE)
data2 <- data2 %>% mutate(cell = replace_na(cell, "HDAC2")) %>% filter(type != "wt_jq12") %>% filter(!str_detect(type, "^wt_ms275"))
ggplot(data2, aes(x=type, y=count, color = cell, group = cell)) +
  geom_point() + stat_smooth(se = FALSE, method = "loess") + scale_y_log10()

data3 <- plotCounts(dds, "ENSG00000171720",
                   intgroup = c("type", "cell"), returnData = TRUE)
data3 <- data3 %>% mutate(cell = replace_na(cell, "HDAC3")) %>% filter(type != "wt_jq12") %>% filter(!str_detect(type, "^wt_ms275"))
ggplot(data3, aes(x=type, y=count, color = cell, group = cell)) +
  geom_point() + stat_smooth(se = FALSE, method = "loess") + scale_y_log10()

data8 <- plotCounts(dds, "ENSG00000147099",
                   intgroup = c("type", "cell"), returnData = TRUE)
data8 <- data8 %>% mutate(cell = replace_na(cell, "HDAC8")) %>% filter(type != "wt_jq12") %>% filter(!str_detect(type, "^wt_ms275"))
ggplot(data8, aes(x=type, y=count, color = cell, group = cell)) +
  geom_point() + stat_smooth(se = FALSE, method = "loess") + scale_y_log10()

```

```{r session_info}
sessionInfo()
```

